Reproducibility of SNP calling in multiple sequencing runs from single tumors

### 2014-10-29 PAPER/paper_d1.tex
revised methods, added table

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

@@

For each sample, we analyzed each technical replicate using the same pipeline (see Methods), and then compared the WGA replicate to the WGS replicate, to see the extent of the similarity between the two. We first looked at the similarity between the two replicates, by number of mutations (Figure 1) and percent overlap partcular mutations (Figure 2) between replicates. We found that overall, the number of mutations found in one replicate correlated strongly with the number of mutations found in the other, though as expected the WGA replicate, with the additional amplification step, had slightly more mutations (Figure 1). We found that the overlap between the two samples, caclulated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was mostly between 20-40\% in WGA replicates and mostly between 40-60\% in WGS replicates. As expected the distribution is narrower and taller in the WGS replicates, as a greater percentage of those samples are likley "real" mutations and so present in the overlap (Figure 2).

% Experiment: number of mutations (figure 3)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
Different cancers mutate at different rates; from some pediatric cancers that arise from single and double hit mutations \cite{pediatric}, to tumors with a mutator phenotype, which usually results from errors in DNA repair pathways \cite{mutator}. GBM specifically is thought to have a relatively low mutation rate \cite{GBM_mut_rate}, and while the majority of our samples had low mutation frequencies in line with this thoery, several samples also had mutation frequencies an order of magnitude greater. One possible explanation is a degraded DNA sample, or bad data. If this were the case, we would expect the percentage of the overlap between replicates to decrease with the overall number of putative mutations. We found no correlation between the number of putatve mutations and the percentage of the sample that was overlapping (Figure 3).

@@

\begin{itemize}
	\item Removes putative SNPs with GATK quality scores less than 40 (as part of the GATK processing, with indel realignment and base recalibration)
	\item Removes ... I forget
	\item Removes putative SNPs with SomaticSniper VAQ scores less than or equal to 20 (VAQ)
	\item Removes putative SNPs that are identified as loss of heterozygosity (LOH)
	\item Removes putative SNPs located within a 10 bp window of any other putative SNP
	\item Removes putative SNPs located within a 10 bp window of indels
	\item Removes putative SNPs that overlap with dbSNP coverage
	\item Removes putative SNPs if, in the tumor data, the percentage of reads covering the site with the alternate allele is less than 10%
\end{itemize}

% Experiment: which filters do what (figures 4 and 5) % Experiment: those two filters are different (figures 6 and 7)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
Large-scale DNA sequencing produces a significant number of errors, both predictable and not \cite{sequence_error}. The best way to validate errors is and remains PCR, but in many cases this is not possible, and having a faster, computational method to "validate" mutations signficantly speeds up genomics research. To this end, one can employ SNP-calling software, and can filter out mutations with specific characteristics. We used the software SomaticSniper \cite{SomaticSniper} and several custom filters to remove from our raw sample putative mutations of with the following traits: (i) a variant allele quality (VAQ), as calculated by SomaticSniper, less than 20, (ii) Loss of heterozygosity (LOH), (iii) within 10 bp of another putative SNP, (iv) within 10 bp of a putative indel, (v) that overlaped wth common human SNPs in dbSNP \cite{dbSNP}, (vi) when the number of reads covering the alternate allele in the tumor sample was less than 10% of the total reads in the tumor sample, (vii) when the Somatic Score (as calculated by SomaticSniper) was less than 40, and (viii) when the quality score (as calculated by GATK) was less than 40. Each of these filters is a common way to computationally clean up data. We first looked at the number of putative mutations per sample in the filtered data and the overlap between replicates of the filtered data. We found only 0-14 putative mutations per sample (average 3) in the filtered data, and a filtered overlap between replicates of 0-2 (average 0). (I don't have a figure for this. Should I?)

Since using all filters destoryed our signal completely, we next looked at the effects of individual filters. Two filters removed nothing. Figure 4 shows the number of putative SNPs removed by each of the remaining 6 filters in all 110 samples (55 WGS and 55 WGA) (Figure 4). We found that different filters removed different numbers of mutations, and that the lion's share of mutations are removed by the VAQ filter and the LOH filter. Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of other mutations, also removed around 50-100 mutatioins each per sample (Figure 4). We next asked how many of the mutations removed by each filter are mutations present in both technical replicates? To answer this, we looked at the percent of the overla that was filtered by each of the six filters. We found that the three filters removeing overlap with dbSNP and other putative mutations removed almost none or none of the overlap, despite removing 50-100 putative mutations on average (Figure 5). In contrast, the VAQ filter (secific to SomaticSniper) and the LOH filter each removed about half of the overlap between replicates (Figure 5). Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different filters. We

%% this is figures 6 and 7, which need to be added to the document and explained now...

@@

All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Gliobastoma multiforme (GBM) data set \cite{TCGA-GBM}. We downloaded raw reads in .bam files from 55 patients using CGHub \cite{CGHub}. For each patient, data consisted of one set of reads taken from blood DNA, and two sets of reads taken from tumor DNA. In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library. On samples (WGS) did not include this step, and on (WGA) did). We developed a pipeline to align all reads to HG19. This pipeline used picard \cite{picard} to generate fastq files from bams aligned to HG18, BWA \cite{BWA} for alignment, SamTools \cite{SAMtools} to clean the data according to accepted standards \cite{seq-standards}, and GATK \cite{GATK} for initial SNP-calling and indel processing. These processes were strung together using custom python code. We used SomaticSniper \cite{SomaticSniper} to align each tumor sequence with it's corresponding blood sequence, then filtered the SomaticSniper data using 6 custom filters. These removed putative SNPs (i) with a SomaticScore less than 40, (ii) within 10 bp from a high-quality indel, (iii) within 10 bp of another putative SNP, (iv) with a variant allele quality less than 20, (v) with a dbSNP \cite{dbSNP} identifier and fewer than 10x coverage, and (vi) that were loss of heterozygosity. All custom code is available in a github repository, and is available upon request.

\subsection{Analysis}

We used custom python scripts, available in the above github repository, to perform simple calculations and data operations. We used R \cite{Rsoftware} to do statistics and generate figures, and this code is also stored in a github repository and available upon request.

@@

%E) (optional) End with a paragraph that again summarizes your research and highlights the most important findings
We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50% of putative SNPs in whole exome sequencing and about 30% in whole genome sequencing. This effect remains even at a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others. While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six strictly computational filters on teh putative mutaions in each technical replicate and the overlap between them, for all 55 samples. We found that some filters remove principally those mutaions found in one sample or the other, while other filers remove primarily those in the overlap. We suggest using just the first set in the future, when working with strictly computational data.

\section{Figures}

>>>>

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
\usepackage{lscape}
% These packages are all incorporated in the memoir class to one degree or another...

@@

For each sample, we called mutations in both technical replicates and the pateint's blood sample using the same computational pipeline (see Figure 1 and Methods): we used CGHub to download TCGA bamfiles aligned to hg18, re-generated teh fastq files with picard, re-aligned the fastq files to hg19 with bwa, performed indel re-alignment and base recalibration with GATK, and finally called somatic mutaitons with SomaticSniper. We then analyzed the list of putative SNPs generated for each replicate by SomaticSniper, as well as the intersection, or overlap, of the two lists. We found that the number of mutations found in one replicate correlated strongly with the number of mutations found in the other, though as expected the WGA replicate, with the additional amplification step, had slightly more mutations (Figure 3). We further found that the overlap between the two samples, caclulated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, mostly between 20-40\% in WGA replicates and mostly between 40-60\% in WGS replicates. As expected the distribution is narrower and taller in the WGS replicates, as a greater percentage of those samples are likley true positives and so more likely to be present in the overlap (Figure 4).

% Experiment: number of mutations (figure 3)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
Different cancers mutate at different rates; from some pediatric cancers that arise from single and double hit mutations \cite{pediatric}, to tumors with a mutator phenotype, which usually results from errors in DNA repair pathways \cite{mutator}. GBM specifically is thought to have a relatively low mutation rate \cite{GBM_mut_rate}, and while the majority of our samples had low mutation frequencies in line with this thoery, several samples also had mutation frequencies an order of magnitude greater. One possible explanation is a degraded DNA sample, or bad data. If this were the case, we would expect the percentage of the overlap between replicates to decrease with the overall number of putative mutations. We found no correlation between the number of putatve mutations and the percentage of the sample that was overlapping (Figure 5). That the the percentage overlap doesn not vary with the number of putative mutaions in either replicate suggests that some samples may simply have a higher mutation rate than others, or indeed than is generally supposed in GBM.

@@

% Experiment: which filters do what (figures 4 and 5) % Experiment: those two filters are different (figures 6 and 7)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
In the normal case lacking sequencing replicates, scientists need other ways to distinguish true and false positives. While the generally accepted standard is PCR, several attempts have been made at computational algorithms that can distinguish somatic mutations from germ line mutations and sequencing error. These platforms typically identify false positives in two ways, (i) first by probablalistic caclulations based on features of the data, and (ii) by features of the mtuations themselves. In this paper, we used GATK for base calibration and indel realignment, and SomaticSniper for somatic mutation calling. Each of these platforms has it's own filters of the first type:

\begin{itemize}
	\item Removes putative SNPs with GATK quality scores less than 40 (as part of the GATK processing, with indel realignment and base recalibration)
	\item Removes putative SNPs with a SomaticScore less than 40 (SS)
	\item Removes putative SNPs with SomaticSniper Varaint Allele Quality scores less than or equal to 20 (VAQ)
\end{itemize}

We additionally employed five filters of the second type, all of which have some support in the computational literature:

\begin{itemize}
	\item Removes putative SNPs that are identified as loss of heterozygosity (LOH)
	\item Removes putative SNPs located within a 10 bp window of any other putative SNP
	\item Removes putative SNPs located within a 10 bp window of indels
	\item Removes putative SNPs that overlap with dbSNP coverage
	\item Removes putative SNPs if, in the tumor data, the percentage of reads covering the site with the alternate allele is less than 10%
\end{itemize}

Our initial question was, does filtering the data in this way increase of decrease the percentage of the sample that is overlapping between the two technical replicates? Alternatively stated, does filtering seem to increase the percentage of probable true positives, or not? We found that after running all eight filters on the data, the number of putative mutaions per preplicate decreases from 50-10,000 to 0-14 (average 3). The size of the overlap drops to 0-2 per sample, with 0 as the mode, so the overall percentage also decreases. We conclude that running all the filters on the data, in the absence of any other verification method, is counterproductive, becuase it seems to remove all of the signal (as well as the noise).

Together, the eight filters we ran drew equally from those putative SNPs found in either replicate and those found in both. We next asked if this was true individually as well, or if different filters acted differently. We did not analyze data on the first two filers, those removing putative SNPs with GATK quality scores below 40 and Somatic Scores below 40, because those filters are part of the Somatic Sniper platform that calls somatic variants, and so none of these putative SNPs were in the unfiltered sample above. We did look at individual data of the other six filters. We first looked at the total number of putative SNPs removed by each filter (figure 6). We found that different filters removed different numbers of mutations, and that the lion's share of mutations are removed by the VAQ and LOH filters, each of which remove 200 or more putative SNPs on average. Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of other mutations, also removed around 50-100 mutatioins each per sample (Figure 6). The final filter, which removed putative SNPs with less than 10% coverage of the alternate allele, removed 0-1 putative SNPs on average, but did on occasion remove as many as 50.

We next asked how many of the mutations removed by each filter are mutations present in the overlap between technical replicates, and how many are in just one sample? Put differently, what percentage of putative SNPs removed by a given filter are in the category more likley to be true positives (overlap), versus the category less likely to be true positives (only one replicate). To answer this, we graphed the percent of the overlap only (per sample) that was filtered by each of the six filters (Figure 7). We found that the three filters removeing overlap with dbSNP and other putative mutations removed almost none or none of the overlap, despite removing 50-100 putative mutations on average. In contrast, the VAQ filter (secific to SomaticSniper) and the LOH filter each removed about half of the overlap between replicates (Figure 7). Thus, our evidence suggests that the filters removeing overlap with dbSNP and putative mutaitons near other putative mutaions are preferentially removing false positives, but the filters removing low VAQ and LOH are less discriminatory.

%% this is figures 6 and 7, which need to be added to the document and explained now...
Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different putative SNPs in each sample. We plotted the percent of the overlap filtered out by the LOH filter against the number of putative mutations in the overlap (figure 8), and the percent of the overlap filtered out by the VAQ filter against the number of putative mutations in the overlap (figure 9). We found opposite trends: the percentage filtered out by VAQ seems to decrease, and the percentage filtered out by LOH seems to increase, with increasing length of overlap. We conclude that the LOH and VAQ filters are not removing the same putative mutations.

@@

All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Gliobastoma multiforme (GBM) data set \cite{TCGA-GBM}. We downloaded .bam files for 55 patients using CGHub \cite{CGHub}. For each patient, data consisted of one bamfile taken from blood DNA, and two bamfiles from tumor DNA, one for each technical replicate. In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library. We developed a pipeline to align all reads to HG19. This connected the pipeline using custom python code (available in a git repository upon request), to include the steps outlined in Figure 1, and given in more detail in Table 1.

\subsection{Filtering and data analysis}

After generating a VCF file will all of the putative somatic variants for each replicate of each sample, we filtered the putative SNPs according to the eight filters described avove using a combination of command line options and custom python code. The GATK quality score and SomaticScore filters were both done with flags when we called SomaticSniper, as shown above (-q 40 -Q 40), so that SomaticSniper returned no putative SNPs that did not pass these tests. SomaticSniper also identified, but did not exclude, LOH mutations; we wrote a custom script to remove them later. We coded custom python scripts for the other five filters, which are in a git repository and available upon request. We also wrote custom scripts to determine the overlap between samples, also available in the git repository upon request.

We plotted all data and did all statistics with standard plotting and statictics functions in R \cite{Rsoftware}. This code is also available upon request.

@@

%E) (optional) End with a paragraph that again summarizes your research and highlights the most important findings
We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNPs in whole exome sequencing and about 30\% in whole genome sequencing. This effect remains even at a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others. While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six strictly computational filters on the putative mutaions in each technical replicate and the overlap between them, for all 55 samples. We found that some filters remove principally those mutaions found in one sample or the other, while other filers remove primarily those in the overlap. We suggest using just the first set in the future, when working with strictly computational data.

\section{Tables and Figures}

\begin{landscape}
\begin{table}
\begin{tabular}{ p{2.5cm} p{4cm} p{14cm} }
	software & purpose & command\\
	\hline
	picard & regenerate fastq files from bamfile aligned to hg18 & java -d64 -Xmx4g -jar SamToFastq.jar I=\$pfx.bam F=\$pfx.1.fastq F2=\$pfx.2.fastq 2$>$\&1 \\
	bwa & align fastq files to hg19 & bwa aln -q 30 -t 8 \$hgReference \$fastq $>$ \$fastq.aln.sai \\
	bwa, samtools & convert aligned fastq files into new bamfile & bwa sampe -a 600 -P -r "\$RG" \$hgReference \$fastq1.aln.sai \$fastq2.aln.sai \$fastq1 \$fastq2 | samtools view -bSh -o \$outprefix.bam - \\
	samtools & sort and index new bamfile & samtools sort -@ 16 \$outprefix.bam \$outprefix.sorted 2, samtools index \$outprefix.sorted.bam 2 \\
	samtools & remove duplicate reads from bamfiles & samtools rmdup ../\$tumorpfx/\$tumorpfx.out.sorted.bam \$tumorpfx.dedup.bam \\
	GATK & indel realignment & java -d64 -jar \$gatkJar -R \$hgReference -T IndelRealigner -rf BadCigar -I \$tumorpfx.dedup.bam -known \$G1000.Mills -known \$G1000.Phase1.Indels -targetIntervals \$tumorpfx.intervals -o \$tumorpfx.realn.bam \\
	GATK & base recalibration & java -d64 -jar \$gatkJar -nct 8 -T BaseRecalibrator -rf BadCigar -I \$tumorpfx.realn.bam -R \$hgReference -knownSites \$dbSNP -o \$tumorpfx.recal.grp \\
	samtools & index recalibrated bamfile & samtools index \$tumorpfx.realn.recal.bam \\
	SomaticSniper & call somatic mutations, generate VCF & bam-somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f \$hgReference \$tumorbam \$normalbam \$tumorpfx.SS.vcf \\
\end{tabular}
\caption{\textbf{Data processing pipeline.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it. The rows are in order of use.}
\end{table}
\end{landscape}
